{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-18 paper/si.tex
memory-one -> memory-1 (for consistency with main text) + fix typos

\noindent
This document provides further details on our methods and derivations, and it contains additional simulation results. 
Section~\ref{section:model} summarizes the model. 
In particular, we provide further details on our implementation of the evolutionary dynamics, and our use of the rare-mutation limit. 
In Section~\ref{section:analyticalresults}, we derive analytical results for the various settings we consider.
These settings differ in what kind of payoff information individuals take into account when updating their strategies. 
In the perfect-memory setting, individuals take into account all their interactions against all co-players.
In the limited-memory setting, they only consider the very last round of their very last interaction.  
In addition, we describe several model extensions in which the amount of information taken into account is in between these two extremes. 
Finally, Section~\ref{section:furthersimulations} presents further simulation results. 
In particular, we confirm that our main results continue to hold ({\it i}) when mutations are no longer rare, and  ({\it ii}) when players use memory-one strategies instead of reactive strategies.

@@

We consider this limit of rare-mutation throughout the main text. 
The respective process is summarised by Algorithm~\ref{algorithm:pairwise_comparison}.
In Section~\ref{section:furthersimulations}, we present additional simulation results to show that our qualitative results continue to hold when the mutation rate is strictly bounded away from zero.

@@

In the following, we discuss our different memory settings in more detail. 
We discuss six cases explicitly. In these cases, updating occurs
({\it i})~based on average payoffs based on all interactions (perfect memory), 
({\it ii})~based on the last round of one interaction (limited memory), 
({\it iii})~based on the last round of two interactions,
({\it iv})~based on the last two rounds of one interaction,
({\it v})~based on the last two rounds of two interactions, and
({\it v})~based on the average payoff of one interaction.
In each case, we assumet there are only two strategies present in the population (a resident and a mutant strategy). 
We first derive how likely it is that a learner (of any type) assigns a given payoff $\pi_\learner$ to itself, and a payoff of $\pi_\rolemodel$ to the role model. 
This allows us to derive explicit expressions for $\lambda^-_k/\lambda^+_k$, and hence for the mutant's fixation probability according to Eq.~\eqref{eq:fixation_probability}.
Based on these expressions we can characterize under which conditions cooperation is stochastically stable.

@@

\subsection{Simulations for memory-one strategies}\label{section:memory_one}

So far, we have assumed that individuals adopt reactive strategies.
To demonstrate that our results hold for more complex strategy sets, 
in the following we present results when players can choose among all memory-one strategies~\citep{sigmund2010calculus}. 
Players with these strategies consider the full outcome of the
previous round to decide on their next action. 
There are four possible outcomes in each
round,  \((C, C), (C, D), (D, C), (D, D)\). 
A memory-one strategy \(\strategy\) can thus be
written as a five-dimensional vector \(\strategy=(y, p_1, p_2, p_3, p_4)\). 
As with reactive strategies, the entry \(y\) is the probability that the strategy opens with a cooperation.
The other entries \(p_1\), \(p_2\), \(p_3\), \(p_4\) are the probabilities that the strategy
cooperates in all subsequent rounds, depending on the outcome of the previous round.
Every reactive strategy $(y,p,q)$ can be written as a memory-one strategy, by using the memory-one representation $(y,p,q,p,q)$.
Conversely, there are memory-one strategies that cannot be represented by a reactive strategy, such as the strategy \texttt{WSLS}$=(1,1,0,0,1)$. 
Therefore, the set of memory-one strategies is a strict superset of the reactive strategies.

To explore the dynamics among memory-one strategies, we perform four separate simulations, for both perfect and limited memory, and for two different benefit values. 
Supplementary~Fig.~\ref{fig:memory_one_low_benefit} shows the results for a comparably low benefit of cooperation; 
Supplementary~Fig.~\ref{fig:memory_one_high_benefit} shows the corresponding results for a large benefit. 
These simulations suggest that even when individuals
are allowed to use memory-one strategies, there is more cooperation when players update their strategies based on perfect payoff memory.

@@

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/memory_one_results_low_benefit.pdf}
  \caption{\textbf{Evolution of memory-one strategies when the benefit of cooperation is comparably small.}
  We perform two independent simulations. In one simulation, individuals use
  expected payoffs. 
  In the other, they update their strategy based on the payoff they received in the last round of their last interaction. 
  We run each simulation for \(T = 10^8\) time steps.
  For each time step, we record the current resident population, which is
  now of the form  \((y, p_1, p_2, p_3, p_4)\). In the left panel, we report the
  cooperation rates for each simulation. 
  For memory-one strategies, like for reactive strategies, the use of expected payoffs results in more cooperation. 
  The right panel reports the most abundant strategy of each
  simulation. Abundance is the number of mutants a strategy can repel before
  being invaded. The most abundant strategies have some similarities,
  namely, \(p_1 \approx 1\), \(p_3 \approx 0\) and \(p_4 > \frac{1}{2}\). 
  Parameters: \(N =100, c=1, b=3, \beta=1, \delta=0.999\).}\label{fig:memory_one_low_benefit}
\end{figure}

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/memory_one_results_high_benefit.pdf}
  \caption{\textbf{Evolution of memory-one strategies when the benefit is high.}
  We consider a similar setup as in Supplementary~Fig.~\ref{fig:memory_one_low_benefit}, but now using a larger benefit value $b\!=\!10$.}\label{fig:memory_one_high_benefit}
\end{figure}

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/cooperation_rate_over_b_and_beta_memory_one.pdf}
  \caption{{\bf Evolution of direct reciprocity for different parameter values for memory-one strategies.}
  To further explore the robustness of our results in the case of memory-one
  strategies, we run simulations for different benefit values (left panels, A)
  and for different selection strengths (right panels, B). These simulations
  suggest that perfect payoff memory consistently leads to more cooperation and
  more generosity. Unless explicitly varied, the parameters of the simulation
  are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}
  \label{fig:cooperation_rate_over_benefit_and_beta_memory_one}
\end{figure}

>>>>

\noindent
This document provides further details on our methods and derivations, and it contains additional simulation results. 
Section~\ref{section:model} summarizes the model. 
In particular, we provide further details on our implementation of the evolutionary dynamics, and our use of the rare-mutation limit. 
In Section~\ref{section:analyticalresults}, we derive analytical results for the various settings we consider.
These settings differ in what kind of payoff information individuals take into account when updating their strategies. 
In the perfect-memory setting, individuals take into account all their interactions against all co-players.
In the limited-memory setting, they only consider the very last round of their very last interaction.  
In addition, we describe several model extensions in which the amount of information taken into account is in between these two extremes. 
Finally, Section~\ref{section:furthersimulations} presents further simulation results. 
In particular, we confirm that our main results continue to hold ({\it i}) when mutations are no longer rare, and  ({\it ii}) when players use memory-1 strategies instead of reactive strategies.

@@

We consider this limit of rare-mutation throughout the main text. 
The respective process is summarized by Algorithm~\ref{algorithm:pairwise_comparison}.
In Section~\ref{section:furthersimulations}, we present additional simulation results to show that our qualitative results continue to hold when the mutation rate is strictly bounded away from zero.

@@

In the following, we discuss our different memory settings in more detail. 
We discuss six cases explicitly. In these cases, updating occurs
({\it i})~based on average payoffs based on all interactions (perfect memory), 
({\it ii})~based on the last round of one interaction (limited memory), 
({\it iii})~based on the last round of two interactions,
({\it iv})~based on the last two rounds of one interaction,
({\it v})~based on the last two rounds of two interactions, and
({\it v})~based on the average payoff of one interaction.
In each case, we assume there are only two strategies present in the population (a resident and a mutant strategy). 
We first derive how likely it is that a learner (of any type) assigns a given payoff $\pi_\learner$ to itself, and a payoff of $\pi_\rolemodel$ to the role model. 
This allows us to derive explicit expressions for $\lambda^-_k/\lambda^+_k$, and hence for the mutant's fixation probability according to Eq.~\eqref{eq:fixation_probability}.
Based on these expressions we can characterize under which conditions cooperation is stochastically stable.

@@

\subsection{Simulations for memory-1 strategies}\label{section:memory_one}

So far, we have assumed that individuals adopt reactive strategies.
To demonstrate that our results hold for more complex strategy sets, 
in the following we present results when players can choose among all memory-1 strategies~\citep{sigmund2010calculus}. 
Players with these strategies consider the full outcome of the
previous round to decide on their next action. 
There are four possible outcomes in each
round,  \((C, C), (C, D), (D, C), (D, D)\). 
A memory-1 strategy \(\strategy\) can thus be
written as a five-dimensional vector \(\strategy=(y, p_1, p_2, p_3, p_4)\). 
As with reactive strategies, the entry \(y\) is the probability that the strategy opens with a cooperation.
The other entries \(p_1\), \(p_2\), \(p_3\), \(p_4\) are the probabilities that the strategy
cooperates in all subsequent rounds, depending on the outcome of the previous round.
Every reactive strategy $(y,p,q)$ can be written as a memory-1 strategy, by using the memory-1 representation $(y,p,q,p,q)$.
Conversely, there are memory-1 strategies that cannot be represented by a reactive strategy, such as the strategy \texttt{WSLS}$=(1,1,0,0,1)$. 
Therefore, the set of memory-1 strategies is a strict superset of the reactive strategies.

To explore the dynamics among memory-1 strategies, we perform four separate simulations, for both perfect and limited memory, and for two different benefit values. 
Supplementary~Fig.~\ref{fig:memory_one_low_benefit} shows the results for a comparably low benefit of cooperation; 
Supplementary~Fig.~\ref{fig:memory_one_high_benefit} shows the corresponding results for a large benefit. 
These simulations suggest that even when individuals
are allowed to use memory-1 strategies, there is more cooperation when players update their strategies based on perfect payoff memory.

@@

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/memory_one_results_low_benefit.pdf}
  \caption{\textbf{Evolution of memory-1 strategies when the benefit of cooperation is comparably small.}
  We perform two independent simulations. In one simulation, individuals use
  expected payoffs. 
  In the other, they update their strategy based on the payoff they received in the last round of their last interaction. 
  We run each simulation for \(T = 10^8\) time steps.
  For each time step, we record the current resident population, which is
  now of the form  \((y, p_1, p_2, p_3, p_4)\). In the left panel, we report the
  cooperation rates for each simulation. 
  For memory-1 strategies, like for reactive strategies, the use of expected payoffs results in more cooperation. 
  The right panel reports the most abundant strategy of each
  simulation. Abundance is the number of mutants a strategy can repel before
  being invaded. The most abundant strategies have some similarities,
  namely, \(p_1 \approx 1\), \(p_3 \approx 0\) and \(p_4 > \frac{1}{2}\). 
  Parameters: \(N =100, c=1, b=3, \beta=1, \delta=0.999\).}\label{fig:memory_one_low_benefit}
\end{figure}

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/memory_one_results_high_benefit.pdf}
  \caption{\textbf{Evolution of memory-1 strategies when the benefit is high.}
  We consider a similar setup as in Supplementary~Fig.~\ref{fig:memory_one_low_benefit}, but now using a larger benefit value $b\!=\!10$.}\label{fig:memory_one_high_benefit}
\end{figure}

\begin{figure}[t!]
  \centering 
  \includegraphics[width=0.8\textwidth]{static/cooperation_rate_over_b_and_beta_memory_one.pdf}
  \caption{{\bf Evolution of direct reciprocity for different parameter values for memory-1 strategies.}
  To further explore the robustness of our results in the case of memory-1
  strategies, we run simulations for different benefit values (left panels, A)
  and for different selection strengths (right panels, B). These simulations
  suggest that perfect payoff memory consistently leads to more cooperation and
  more generosity. Unless explicitly varied, the parameters of the simulation
  are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}
  \label{fig:cooperation_rate_over_benefit_and_beta_memory_one}
\end{figure}
